A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus \\ (Draft)

### 2014-08-05 Papers/NaturalDeduction/GodProof-ND.tex
Improve introduction and conclusion of ND paper

@@

\title{G\"{o}del's Ontological Proof of God's Existence (Draft)}

@@

ToDo: Do also Scott's and G\"{o}del's proofs.

The first ontological argument for the existence of God was written by St. Anselm (1033-1109). The argument considers a maximally conceivable being. This being must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also has the property of existence. The main critique of this argument is that we do not know whether the concept maximal conceivable being in fact designates anything or if it is inconsistent, like a round square \citep{fitting}[p.134].  The question is if the argument is sound.

As Bertrand Russell has pointed out the definition of maximal allows us to define properties, like having boots, which the maximal being then also must have. Kant, on the other hand argued against the ontological argument on the basis that existence is not an analytic property \citep{kant}. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. All that we can say is that if God exists, then God necessarily exists.

St. Anselm's proof was processed by Descartes and Leibniz. Leibniz identified the critical point of the argument as establishing the possible existence of God. Leibniz gave an argument for that the properties of God, the perfections, are compatible. This implies that it is possible to have all perfections at once and therefore the existence of a maximal being with all these properties is possible.

G\"odel continued this argument by defining the properties that must hold for these perfections and giving axioms from which the existence of God can be derived. The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The formal argument of G\"odel is based on Leibniz proof, which in turn is based on Descartes proof. These proofs have two parts; a proof that if God's existence is possible, then it is necessary and a proof that God's existence is in fact possible. The deductive system used in the formal proof is a system for modal logic extended with second-order quantification (or possibly third-order quantification), because the axioms and definitions required for the concept of positive properties or perfections involve quantification over properties.

G\"odel treated positive properties as not just atomic properties, like Leibniz, but also consisting of collections of these properties, which by Leibniz argument are compatible or possible.\citep{fitting}[p.139.] G\"odel formulated an axiom stating that the conjunction of any set of positive properties is positive. The property of God-likeness, which defines what a God is, must therefore be an infinite conjunction of all the positive properties. This axiom of the positiveness of conjunctions can be given a third order formulation which is the reason for the claim that third-order quantification may be required \citep{and}. %\citep{fitting}[p. 148.]   
However, this axiom is only used in the proof to establish that the property of God-likeness is itself positive. G\"odel's student Scott, who wrote down a version of G\"odel's proof after G\"odel had confided his work in him, simply chose to assume the positiveness of  God-likeness as an axiom instead.

Nevertheless, both axioms have the unfortunate property of being equivalent to the possibility of God's existence, given the other axioms of G\"odel \citep{sobel2}.  %\citep{fitting}[Section 11.4. Objections.] 
The possibility  God's existence in turn is equivalent to  God's necessary existence as well as  God's existence itself. Thus, we have an axiom which seems to be equivalent to the conclusion.

Another, possibly undesirable, consequence of the formalism is that there can only be one God, which is defined by the axioms. Namely, if one assumes that the theory contains the equality relation then a God has the property of being identical to itself. Since any other god-like being would have at least the same properties it shares the property of being identical to the one God. Thus, we have proved monotheism \citep{sobel2}[Ch. 4, section 3.3.2]. %\citep{fitting}[The proof is Exercise 7.1, p.163].

In \citep{sobel} a concern is raised about the whole system of proof, which is further elaborated on in \citep{sobel2}[Ch. 4,  section 6]. Depending on the properties of the rules for second order quantification allowed in the system, the system may have modal collapse. That is, if no restrictions are placed on the properties over which quantification is allowed, then the system says that for all formulas the formula implies its boxed version.  However, there are solutions to this problem by modification of the axioms. Sobel's own solution to the problem is to treat properties as extensional instead of intentional. Another solution by Anderson %(1990) 
is to require that Godlike beings have positive properties necessarily. A summary of these arguments can be found in Fitting \citep{fitting}[Sections 11.9 and 11.10].

\section{Natural Deduction}

@@

\section{Possibly, God Exists}

@@

\section{Being God is an essence of any God}

@@

\section{If God's existence is possible, it is necessary}

@@

\section{God exists}

@@

\section{Necessarily, God exists}

@@

\section{God exists}\label{GodsE}
This section is superfluous.

@@

The axioms of the calculus imply the following collapse of modalities:

@@

This collapse of the system was first proved in \citep{sobel} and the argument is also given in \citep{fitting}[ch. 11, sections 8-11] who provides summaries of two solutions by the modifications of the axioms. The argument may require that one doesn't instantiate the second order quantifier with $A$ directly but with $\lambda x.A$ to make $A$ a property. But still we have $(\lambda x.A)(y)\equiv A$ because $A$ is constant.

The question arises what kind of second order quantification is allowed. The instantiations that are used in our proof are instantiated with predicate variables that later work as eigenvariables for the second order universal introduction. In the proof of corollary \ref{C1} we instantiate with $G$ and in the proof of lemma \ref{L1}. we instantiate with $G$ and $E$. %In Sobel on G\"{o}del's ontological proof by Robert E. Koons this question is raised.

One can also discuss if a second order modal logic is sound and complete. With Henkin models, which restrict the second order quantification with comprehension, we have that normal second order logic is sound and complete. There is a completeness proof for a second order modal logic found in A completeness theorem in second order modal logic by Cocchiarella.

>>>>

\title{A Variant of G\"{o}del's Ontological Proof \\
in a Natural Deduction Calculus \\
(Draft)}

@@

@@

Ontological arguments for the existence of God can be traced back at least to St. Anselm (1033-1109). His argument considered a maximally conceivable being, who must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also has the property of existence.

A major critique of this argument is that we do not know whether the concept maximal conceivable being in fact designates anything or if it is inconsistent, like a round square \citep{fitting}[p.134].  
%The question is if the argument is sound. 
As Bertrand Russell pointed out, the definition of maximal allows us to define properties, like having boots, which the maximal being then also must have \citep{citation needed}.

Kant argued against the ontological argument on the basis that existence is not an analytic property \citep{kant}. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. All that we can say is that if God exists, then God necessarily exists \citep{citation needed}.

St. Anselm's argument was elaborated further by Descartes and Leibniz. Leibniz identified that establishing the possible existence of God is a critical missing step in St. Anselm's argument. To fill this gap, he argued that the properties of God, the perfections, are compatible. This implies that it is possible to have all perfections at once and therefore the existence of a maximal being with all these properties is possible.

G\"odel studied Leibniz's work \citep{Adams} and brought his ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being having all positive properties.

G\"odel's notion of positive property and Leibniz's notion of perfection are not exactly the same \citep{Adams}. On a technical level, the main distinction seems to be that G\"odel's positive properties are not just atomic properties, like Leibniz's perfections, but can also consist of complex combinations of atomic properties \citep{fitting}[p.139.] (TODO: Check this claim). In particular, one of G\"odel's axioms states that the conjunction of any set of positive properties is positive. And from this axiom, it follows immediately that the property of being God-like is positive. While this step is intuitively and informally clear, it is not easily formalizable in a typical logical calculus, because it requires inferring that being god-like is a (possibly infinite) conjuntion of positive properties, while it has only been defined as the property holding for individuals who have all positive properties. This interplay between universal quantification (in the definition of a God-like being) and conjunction (in G\"odel's mentioned axiom) is a technical difficulty that probably explains why, starting with Scott \cite{scott}, to whom G\"odel confided his manuscript, this axiom of G\"odel's has been replaced by another one that simply assumes the positivity of the property of being God-like. 
%This axiom is unfortunately equivalent to the possibility of God's existence given the other axioms, leading to criticisms that it might be too strong \citep{sobel2}\citep{fitting}[Section 11.4. Objections.].

%The possibility  God's existence in turn is equivalent to  God's necessary existence as well as  God's existence itself. Thus, we have an axiom which seems to be equivalent to the conclusion.

 % This axiom of the positiveness of conjunctions can be given a third order formulation which is the reason for the claim that third-order quantification may be required \citep{and}. %\citep{fitting}[p. 148.]

% Another, possibly undesirable, consequence of the formalism is that there can only be one God, which is defined by the axioms. Namely, if one assumes that the theory contains the equality relation then a God has the property of being identical to itself. Since any other god-like being would have at least the same properties it shares the property of being identical to the one God. Thus, we have proved monotheism \citep{sobel2}[Ch. 4, section 3.3.2]. %\citep{fitting}[The proof is Exercise 7.1, p.163].

This paper presents new natural deduction proofs of (Scott's version of) the lemmas and theorems in G\"odel's ontological argument \cite{Goedel}. In contrast to Scott's version \cite{Scott}, also shown here in natural deduction form, the version presented here is characterized by shorter and more direct derivations.

Section \ref{sec:calculus} defines the higher-order modal natural deduction calculus used to construct the formal proofs. Section \ref{sec:scott} shows Scott's proofs, translated to natural deduction. And section \ref{sec:proof} presents the new proofs.

\section{Natural Deduction}
\label{sec:calculus}

@@

\section{Scott's Proof}

The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The formal argument of G\"odel is based on Leibniz proof, which in turn is based on Descartes proof. These proofs have two parts; a proof that if God's existence is possible, then it is necessary and a proof that God's existence is in fact possible.

\subsection{Possibly, God Exists}

@@

\subsection{Being God is an essence of any God}

@@

\subsection{If God's existence is possible, it is necessary}

@@

\subsection{God exists}

@@

\subsection{Necessarily, God exists}

@@

\subsection{God exists}\label{GodsE}
This section is superfluous.

@@

A major criticism against G\"odel's proof is that its axioms lead to the so-called \emph{modal collapse} \citep{sobel}: it is possible to prove that everything that is the case is so necessarily, and hence actuality, possibility and necessity coincide \citep{sobel2}[Ch. 4,  section 6, theorems 9 and 10]. As this is an undesirable consequence, many solutions to the problem of the modal collapse have been proposed.

Anderson's solution \citep{citation needed} modifies the definitions of God-like being and essence, and eliminates half of an axiom. This not only avoids the modal collapse, but also makes two of G\"odel's five axioms derivable from the others \citep{hajek} under some implicit additional assumptions \citep{fuhrmann}. Another solution involving more substantial modifications is that of Bj{\o}rdal \citep{bjordal,fuhrmann}.

On another track, Fitting has argued that greater care has to be taken with the semantics of higher-order modal logics. Quantified variables may be rigid or flexible; and properties may be treated as intensional or extensional. Making the right choices may prevent the modal collapse \citep{fitting}[Sections 11.9 and 11.10].

Anderson \citep{anderson1990}[p. 292] and Sobel \citep{sobel2}[p. 133] also discuss the idea that the notion of property over which quantification is allowed might be too general and restrictions might be appropriate.

%In his manuscript, G\"odel warned that ``positive means positive in the moral aesthetic sense [\ldots]. Only then [are] the axioms true''.

It is beyond the scope of this paper to analyze these solutions in detail or propose new solutions. The purpose of this section is simply to show natural deduction derivations of the modal collapse, thus confirming that it holds for the axioms used in the previous sections.

@@

% This collapse of the system was first proved in \citep{sobel} and the argument is also given in \citep{fitting}[ch. 11, sections 8-11] who provides summaries of two solutions by the modifications of the axioms. The argument may require that one doesn't instantiate the second order quantifier with $A$ directly but with $\lambda x.A$ to make $A$ a property. But still we have $(\lambda x.A)(y)\equiv A$ because $A$ is constant.

% The question arises what kind of second order quantification is allowed. The instantiations that are used in our proof are instantiated with predicate variables that later work as eigenvariables for the second order universal introduction. In the proof of corollary \ref{C1} we instantiate with $G$ and in the proof of lemma \ref{L1}. we instantiate with $G$ and $E$. %In Sobel on G\"{o}del's ontological proof by Robert E. Koons this question is raised.

% One can also discuss if a second order modal logic is sound and complete. With Henkin models, which restrict the second order quantification with comprehension, we have that normal second order logic is sound and complete. There is a completeness proof for a second order modal logic found in A completeness theorem in second order modal logic by Cocchiarella.

\section{Conclusions}

The proofs of theorem T1 and corollary C1 ($\pos \ex x. G(x)$) do not rely on equality and use axiom A2 ($\all \varphi. \all \psi.[(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)]$) only once. 
Corollary C2 ($\ex x. G(x)$), which is usually regarded as a trivial corollary of main theorem T3 ($\nec \ex x. G(x)$) using the modal logic axiom T, is here derived directly from lemma L1 and corollary C1, not relying on T. The main theorem T3 becomes derivable from C2 by a single application of the necessitation rule. Furthermore, all proofs are done in the modal logic K, except for the proof of corollary C2, which requires one use of the axiom B of modal logic KB.

ToDo: Sobel Anderson on KB: Sobel page 152
